Clustering Beat-Chroma Patterns in a Large Music Database

### 2010-03-23 Paper/Paper.tex
fleshed out/reorganized section 4 a bit more

\section{Visualization} \label{sec:visu}
The goal here is to give an intuition to the reader of what kind of
typical patterns do we get. We also want to have a feel for what the
clusters look like, how many boring sustained notes are encoded in the
codebook, can the codebook describes the complexity of the music space, etc.

Below we present some figures to help the reader with those questions.
Please bear in mind the limited space.

\subsection{Encoding of a Song}
Here we present a song with its two encodings. First, the harmonic patterns
given by the EchoNest. Secondly, each pattern is encoded by its closest
code in the codebook. See Figure \ref{fig:encodesong}.

\subsection{Codebook}
Visualizing codewords in a meaningful way is always a challenge, even
if we have only a one or two hundreds of them.  We present $3$ figures
in this section that should help the reader foster his
intuition.

@@

The 25 most frequently used codewords in the testing set are shown in
Figure \ref{fig:codes1}.  The codewords primarily consist of sustained
notes and simple chords.  Since they are designed to be key-invariant,
specific notes do not appear.  Instead the first 7 codewords
correspond to a single note sustained across two bars (codeword 1),
perfect fifth (codes 2 and 3) and fourth intervals (codes 4 7), and a
major and minor triads (code 6 and 5).  Many of the remaining
codewords correspond to transitions from one chord to another, \eg a
transition from one major triad to another (\ie Cmaj to Fmaj assuming
that chroma 0 is C) in code 8.  Does this transition have a better
name like ``relative x''?

@@

Finally, more related to our particular data, we try to assess how many
sustained patterns are represented in the codebook. We look at
the average variance over time, and we get an estimate of $60\%$ of the
codebook (Figure \ref{fig:code_var}).

@@

\subsection{Cluster}
Once we understand the kind of codebook we create, we naturally want
to visualize the clusters, \ie the pattersn from actual song that
surround a given centroid. Unsurprisingly, real patterns tend to be
noisier than codes. However, centroids seem to capture important aspects
of the patterns. We could compare the centroids with the result of
a lowpass filtering.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.8\columnwidth]{close_patterns_diff}
\end{center}
\caption{\small{First three patterns of figure \ref{fig:cluster}
($2$nd line) presented with the centroid from Figure \ref{fig:code}
($1$st line) and the difference between both ($3$rd line).
}}
\label{fig:cluster_diff}
\end{figure}

>>>>

\section{Visualization} \label{sec:visu}
% The goal here is to give an intuition to the reader of what kind of
% typical patterns do we get. We also want to have a feel for what the
% clusters look like, how many boring sustained notes are encoded in the
% codebook, can the codebook describes the complexity of the music space, etc.

\subsection{Codebook}\label{sec:codebook}

How was this codebook trained?
What do we expect it to capture?

@@

The 25 most frequently used codewords in the test set are shown in
Figure \ref{fig:codes1}.  The codewords primarily consist of sustained
notes and simple chords.  Since they are designed to be key-invariant,
specific notes do not appear.  Instead the first 7 codewords
correspond to a single note sustained across two bars (codeword 1),
perfect fifth (codes 2 and 3) and fourth intervals (codes 4 7), and a
major and minor triads (code 6 and 5).  Many of the remaining
codewords correspond to common transitions from one chord to another,
\eg a transition from one major triad to another (\ie Cmaj
$\rightarrow$ Fmaj assuming that chroma 0 is C) in codes 8 and 10 and
the reverse transition in code (\ie Fmaj $\rightarrow$ Cmaj) in code
22.  Does this transition have a better name like ``relative x''?

@@

Finally, more related to our particular data, we try to assess how many
sustained patterns are represented in the codebook. We look at
the average variance over time, and we get an estimate of $60\%$ of the
codebook (Figure \ref{fig:code_var}).  25\% of the codebook correspond
to very simple sustained patterns similar to the top row of Figure
\ref{fig:codes1}.  How is this different from the next 35\%?

@@

\subsection{Cluster}
Once we understand the kind of codebook we create, we naturally want
to visualize the clusters, \ie the patterns from actual song that
surround a given centroid. Unsurprisingly, real patterns tend to be
noisier than codes. However, centroids seem to capture important aspects
of the patterns. We could compare the centroids with the result of
a lowpass filtering.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.8\columnwidth]{close_patterns_diff}
\end{center}
\caption{\small{First three patterns of figure \ref{fig:cluster}
($2$nd line) presented with the centroid from Figure \ref{fig:code}
($1$st line) and the difference between both ($3$rd line).
}
ronw: does this figure really add all that much?
}
\label{fig:cluster_diff}
\end{figure}

\subsection{Example song encoding}

Figure \ref{fig:encodesong} shows a typical example of encoding a song
using the codebook described in Section \ref{sec:codebook}.  Both the
Echo Nest chroma features for an example song and the corresponding
quantization are displayed.  The quantized representation retains
dominant harmonic structure of the original features, but is smoothed
in time due to the use of 2 bar long codewords.
%
This is consistent with the codebook described in Section
\ref{sec:codebook} which primarily consists of sustained chords across
4 or 8 beats.
